---
title: "Data Containers Homework"
subtitle: "Building a manuscript and its analysis together"
author: "YOUR NAME HERE"
format:
    html:
        self-contained: true
        code-tools: true
---

## The Idea

One of the central habits of reproducible research is that the **words and the code that produce them are written together**. A paragraph of *Methods* is followed immediately by the chunk that carries out what it describes. A statement in *Results* is backed by a value generated in the chunk beside it, and for numbers reported in the running text the value is read straight from a variable in memory using an inline `` `r ` `` expression, never typed by hand.[^ref-inline]

[^ref-inline]: For a fuller treatment of pulling values out of an analysis and citing them inline, see the narrative section [Analysis Output](https://dyerlabteaching.github.io/Data-Containers/narrative.html#analysis-output) and the slide [Benefits](https://dyerlabteaching.github.io/Data-Containers/slides.html#/benefits).

This assignment has you practice that habit on a small analysis of the `iris` data set. Read the worked example, then reproduce its structure for sepal width and petal width.

## A Worked Example

### Methods

> **Data.** We used Edgar Anderson's *iris* measurements, distributed with R as the `iris` data set (Anderson 1935): four floral traits — sepal and petal length and width, in centimeters — recorded on 150 specimens, 50 from each of three species (*Iris setosa*, *I. versicolor*, *I. virginica*). For this example we pulled sepal length and petal length for all 150 specimens, all species pooled, into a pair of vectors.

```{r}
data("iris")
sepal_length <- iris$Sepal.Length
petal_length <- iris$Petal.Length
```

> **Analysis.** We summarized each trait by its mean and tested the association between the two traits with a Pearson product-moment correlation using `cor.test()`. We report the correlation coefficient, the *t*-statistic, and the *p*-value.

```{r}
mean( sepal_length )
mean( petal_length )
fit <- cor.test( sepal_length, petal_length )
fit
```

### Preparing the reporting values

Before writing a word of the Results, spend a moment pulling the numbers you are going to report out of the `fit` object and formatting them. You *could* write each one inline — `` `r format(fit$estimate, digits = 3)` `` — but a paragraph full of `format()` calls is hard to read and harder to revise: every time you reread the sentence you have to step over the code to find the prose. Doing that work in one small chunk, and giving each value a short name, keeps the writing clean.

```{r}
mean_sepal <- format( mean(sepal_length), digits = 3 )
mean_petal <- format( mean(petal_length), digits = 3 )

r <- format( fit$estimate,  digits = 3 )
t <- format( fit$statistic, digits = 3 )
p <- format( fit$p.value,   digits = 3 )
N <- length( sepal_length )
```

### Results

> Across all `r N` specimens, mean sepal length was `r mean_sepal` cm and mean petal length was `r mean_petal` cm. Sepal and petal length were strongly and positively correlated (Pearson correlation, *r* = `r r`, *t* = `r t`, *P* = `r p`).

Now the source of that paragraph reads almost exactly as the rendered sentence will, and if the analysis changes you edit only the chunk above, never the prose.

Here is the raw text that produced it:

```{verbatim}
Across all `r N` specimens, mean sepal length was `r mean_sepal` cm and mean petal length was `r mean_petal` cm. Sepal and petal length were strongly and positively correlated (Pearson correlation, *r* = `r r`, *t* = `r t`, *P* = `r p`).
```

## Your Assignment

Following the worked example, produce a standalone **Methods** and **Results** write-up for the correlation between **sepal width and petal width** across all 150 *iris* specimens.[^ref]

Your document should have:

1. **Methods** — a *Data* paragraph followed by the chunk that pulls the two width vectors out of `iris`, then an *Analysis* paragraph followed by the chunk that takes the mean of each and runs `cor.test()`.
2. **A reporting-values chunk** — `format()` the two means and the correlation's *r*, *t*, and *p*, and give each a short name.
3. **Results** — one paragraph reporting those numbers as bare inline values (`` `r r` ``), with no `format()` in the prose.

Then answer briefly, a sentence or two each:

- Name the container type of `iris`, your width vectors, and `fit`, and say why each suits what it holds.
- Why does `cor.test()` return a list rather than a data frame?

[^ref]: The mechanics — pulling vectors out of a data frame, `cor.test()` output as a list, reporting values inline — are all in the module [narrative](https://dyerlabteaching.github.io/Data-Containers/narrative.html) and [slides](https://dyerlabteaching.github.io/Data-Containers/slides.html).

## What You Turn In

**One file: the rendered, self-contained `.html`.** Not the `.qmd`, and not a copy of this page. Make a *new* Quarto document that contains only your own work — your Methods and Results sections with their chunks, the reporting-values chunk, and your brief answers to the two questions.

The worked example above is a model to imitate, not text to paste into your document.

### Making the document

1. In RStudio: **File ▸ New File ▸ Quarto Document…**
2. In the dialog, enter a title and your name, leave the engine set to **Knitr**, and click **Create**.
3. Delete everything RStudio puts below the header (the "Quarto" boilerplate and its example chunk).
4. Replace the header it generated with the one below.
5. Save the file into your course project folder as `Data-Containers-Homework.qmd`.
6. Write your content, then click **Render** to produce the `.html`. That HTML file is what you submit.

### The full YAML header

```yaml
---
title: "Data Containers Homework"
author: "Your Name"
date: today
abstract: |
  A sentence or two saying what the document does: the data, the question, and what you found.
format:
  html:
    self-contained: true
    code-tools: true
---
```

- **`title` / `author` / `date` / `abstract`** — complete front matter, including a short summary where a scientific paper would put its abstract. `date: today` fills itself in at render time.
- **`format: html`** — render to a single web page I can open in any browser.
- **`self-contained: true`** — bundle every image, style, and script *into* that one `.html` file. Without it the file depends on a sibling `*_files/` folder that never makes it into the upload, and your figures and formatting arrive broken.
- **`code-tools: true`** — adds the "</> Code" button in the top-right that lets a reader view the full source. **This is not optional.** If I cannot see the code behind a wrong answer, I cannot tell whether you were one typo away or lost the plot entirely — which means I cannot give partial credit, point out where it went sideways, or help you fix it. A rendered page with no visible code is graded only on what rendered.

Use this same header layout for every document you turn in this course.

## How This Is Graded

This assignment is scored with the **Online Homework Rubric** on the Canvas site — the same rubric used for every online homework in this course. It covers online access to the published document, front matter, formatting, and whether the document stands on its own; the remaining points come from the assignment-specific work above.
