---
title: "Data Containers Homework"
subtitle: "Building a manuscript and its analysis together"
author: "YOUR NAME HERE"
format:
    html:
        self-contained: true
---

## The Idea

One of the central habits of reproducible research is that the **words and the
code that produces them are written together**. A paragraph of *Methods* is
followed immediately by the chunk that carries out what it describes. A statement
in *Results* is backed by a value, table, or figure generated in the chunk beside
it — and for numbers reported in the running text, the value is read straight
from a variable in memory using an inline `` `r ` `` expression, never typed by
hand.

This assignment has you practice that habit on a small analysis of the `iris`
data set. First read the worked example, then reproduce its structure for a
different pair of variables and extend it.

## A Worked Example

### Methods

> **Data.** We used Edgar Anderson's *iris* measurements, distributed with R as
> the `iris` data set (Anderson 1935): four floral traits — sepal and petal
> length and width, in centimetres — recorded on 150 specimens, 50 from each of
> three species (*Iris setosa*, *I. versicolor*, *I. virginica*). For this
> example we pulled sepal length and petal length for all 150 specimens, all
> species pooled, into a working data frame.

```{r}
data("iris")
morph <- data.frame(
  sepal_length = iris$Sepal.Length,
  petal_length = iris$Petal.Length
)
```

> **Analysis.** We summarised each trait by its mean and tested the association
> between the two traits with a Pearson product-moment correlation using
> `cor.test()`. We report the correlation coefficient, the *t*-statistic, and the
> *p*-value.

```{r}
mean( morph$sepal_length )
mean( morph$petal_length )
fit <- cor.test( morph$sepal_length, morph$petal_length )
fit
```

### Results

> Across all 150 specimens, mean sepal length was
> `r format(mean(morph$sepal_length), digits=3)` cm and mean petal length was
> `r format(mean(morph$petal_length), digits=3)` cm. Sepal and petal length were
> strongly and positively correlated (Pearson correlation, *r* =
> `r format(fit["estimate"], digits=3)`, *t* =
> `r format(fit["statistic"], digits=3)`, *P* =
> `r format(fit["p.value"], digits=3)`).

Look at the source of that paragraph: every number is an inline call to a
variable already computed in the chunk above it. Nothing is hard-typed, so if the
data changed, the paragraph would rewrite itself on the next render.

## Your Assignment

Reproduce this pattern for **sepal width and petal width**, then extend it.

Use **all 150 specimens** for the main analysis — do not filter by species yet
(selective filtering with `filter()` and friends is the next module, Tidyverse).

### 1. Methods

Write **two short paragraphs, each followed immediately by the chunk that
implements it**:

a. A **Data** paragraph saying what you extracted, followed by a chunk that pulls
   `iris$Sepal.Width` and `iris$Petal.Width` (and, for Part 3, `iris$Species`)
   into a data frame of your own.

```{r}
#| label: assignment-data
# Build your data frame here.
```

b. An **Analysis** paragraph describing the summary and the test, followed by a
   chunk that computes the mean of each width column and runs `cor.test()` on the
   two.

```{r}
#| label: assignment-analysis
# Compute the means and run cor.test() here.
```

### 2. Results

Write **one paragraph** reporting the two means and the correlation (*r*, *t*,
and *P*). Report every value with an inline `` `r ` `` expression using
`format(..., digits = 3)`, exactly as in the worked example. This section
contains **no printed chunk output, no tables, and no figures** — just the prose
and its inline values.

### 3. Compare a species pair

Now ask whether the association you found across all three species looks the same
*within* a narrower group.

- Add `iris$Species` to your data frame if you have not already.
- Pick **one pair of species** (for example *versicolor* and *virginica*) and
  subset your data frame to just those rows using logical indexing, e.g.
  `df[ df$species %in% c("versicolor", "virginica"), ]`.
- Run `cor.test()` on sepal width vs petal width **within that pair**.

```{r}
#| label: assignment-pair
# Subset to your species pair and run cor.test() on the subset.
```

Then add **one or two sentences to your Results paragraph** (still inline values
only) that report the within-pair correlation and state whether it is
**stronger or weaker** than the all-species correlation, and by roughly how much.

### 4. Container check

Short answers — these make sure you can name and handle each container.

1. Run `class()`, `length()`, and `is.vector()` on one of your width columns.
2. Use `cbind()` to build a matrix from your two width vectors. Report its
   `dim()`, then show the first row (`M[1, ]`) and the second column (`M[, 2]`).
3. `names(fit)` shows that the object returned by `cor.test()` is a list. In one
   sentence, why does `cor.test()` return a **list** rather than a **data frame**?

```{r}
#| label: assignment-containers
# class(), length(), is.vector(), cbind(), names() here.
```

### 5. Reflection

Two or three sentences each:

1. Name the container type of `iris`, your width data frame, the two mean values,
   and `fit`, and say why each is the right container for what it holds.
2. In this course you were told you will really only use a list in one situation.
   What is it, and does anything in this assignment match that description?
3. When a within-pair correlation differs from the pooled one, what does that tell
   you about combining groups before analysis?

## What You Turn In

A rendered, self-contained HTML file containing: the worked example (unchanged),
your two Methods paragraphs with their chunks, your Results paragraph with inline
values only, your species-pair comparison, and your answers to the container
check and reflection.

## How This Is Graded

| Component | Module Learning Objective |
|---|---|
| Data frame built from `iris` vectors; container check (Part 4) | Apply & Analyze Data Container Structures |
| Analysis chunk; Results paragraph reported entirely with inline references | Integrate Data Containers into Reproducible Workflows |
| Species-pair subset and comparison (Part 3) | Both — data-frame manipulation and reproducible reporting |
| Reflection (Part 5) | The *Analyze* dimension of both objectives |
