---
title: "Narrative Title"
format: html
editor: visual
---

> Data needs to be contained within objects so that we can examine, manipulate, sort, analyze, and communicate about it. In this topic, we examine some of the more basic

## Vectors

```{r}
x <- c(1,2,3)
x
```

```{r}
class(x)
```

```{r}
length(x)
```

```{r}
is.vector( x ) && is.numeric( x )
```

```{r}
1:10 -> y
y
```

```{r}
seq(1,10,by=3)
```

```{r}
seq( 119, 121, length.out = 6)
```

```{r}
x[3]
```

```{r}
x[5]
```

```{r}
x[2] <- 42
x
```

```{r}
x[7] <- 43
x
```

```{r}
x <- 1:5
y <- rpois(5,2)
x
y
```

```{r}
x + y 
```

```{r}
x^y
```

*recycling rule*

```{r}
z <- 1:10
x + z
```

## Matrices

```{r}
X <- matrix( 1:9, nrow=3, ncol=3 )
X
```

```{r}
X[2,2] <- 42
X
```

```{r}
Y <- t(X)
Y
```

```{r}
X + Y
```

```{r}
X * Y
```

```{r}
X %*% Y
```

## Lists

```{r}
lst <- list("A","B",323423.3, pi, TRUE)
lst
```

```{r}
lst[[2]] <- "C"
lst
```

### Named Lists

```{r}
myInfo <- list()
myInfo["First Name"] <- "Rodney"
myInfo["Second Name"] <- "Dyer"
myInfo["Favorite Number"] <- 42
myInfo["Vegitarian"] <- FALSE

```

```{r}
myInfo
```

```{r}
myInfo$Homes <- c("RVA","Oly","SEA")
myInfo
```

```{r}
names(myInfo)
```

### Analysis Output

```{r}
iris.test <- cor.test( iris$Sepal.Length, iris$Petal.Length )
x
```

```{r}
names(iris.test)
```

Because of this, we can easily incorporate the findings of analyses, such as this simple correlation test, and insert the content into the text. All you have to do is address the components of the analysis as in-text r citation. Here is an example where I include the values of:

```{r}
iris.test$estimate
iris.test$statistic
iris.test$p.value
```

Here is an example paragraph (see the raw quarto document to see the formatting).

> There was a significant relationship between sepal and petal length (Pearson Correlation, $\rho =$ `r format(iris.test["estimate"], digits=3)`, $t =$ `r format(iris.test["statistic"], digits=3)`, P = `r format(iris.test["p.value"],digits=3)`).

## Data Frames

```{r}
x <- 1:10
y <- LETTERS[11:20]
z <- sample( c(TRUE,FALSE), size=10, replace=TRUE )
```

```{r}
df <- data.frame( TheNums = x,
                  TheLetters = y,
                  TF = z
                  )
df
```

```{r}
summary( df )
```

```{r}
names( df )
```

```{r}
df$TheLetters
class( df$TheLetters )
```

Indexing, by named column and by `[row,col]` values.

```{r}
df$TheLetters[3]
df[3,2]
```

```{r}
dim( df )
nrow( df )
ncol( df )
```

### Loading Data

```{r}
#| eval: false
library( readr )
```

```{r}
#| echo: false
if( !require(readr)) { 
  install.pacakges("readr")
  library( readr )
}
```

```{r}
url <- "https://raw.githubusercontent.com/DyerlabTeaching/Data-Containers/main/data/arapat.csv?token=GHSAT0AAAAAABYCMZEFK5KI5I2YGMJ6A7PKYYY5LEQ"
```

```{r}
#| message: false
beetles <- read_csv( url )
summary( beetles )
```

```{r}
#| eval: false
library( leaflet )
```

```{r}
#| echo: false
if( !require( leaflet ) ) { 
  install.packages("leaflet")
  library( leaflet)
}
```

```{r}
beetles %>%
  leaflet() %>%
  addProviderTiles(provider = providers$Esri.WorldTerrain) %>%
  addMarkers( ~Longitude, ~Latitude,popup = ~Stratum )
```

## Questions

If you have any questions for me specifically on this topic, please post as an Issue in your repository, otherwise consider posting to the discussion board on [Canvas](https://canvas.vcu.edu).
